Improved state-space models for inference about spatial and temporal variation in abundance

### 2012-09-04 Dail_Mad_ext.tex
bib, intro

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}

@@

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or management actions. State-space
models, are widely used for these purposes, and they allow for
estimating parameters of classical growth models, such as exponential,
logistic, and Gompertz, while accounting for demographic stochasticity
and observation error. Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models for spatially-replicated
time-series data recently proposed by
\citet[Biometrics]{dail_madsen:2011}. We
expand this class of models to accommodate classical growth
models, zero-inflation, and random effects such as observer-specific
detection probabilities. Furthermore, we describe classical and
Bayesian methods of parameter estimation. These new developments will
allow researchers to apply these methods to address important
questions regarding the factors affecting spatial and temporal
variation in abundance. We demonstrate these developments by analyzing
data from the North American Breeding Bird Survey. Code to fit these
models using the R package unmarked and JAGS is also provided.
}

@@

Theoretical ecology requires robust models for testing hypotheses
regarding density dependence and spatial variation in population
dynamics. Such models are also required in applied ecological research
for predicting species' responses to environmental change or
management actions. For example, many researchers are currently
investigating the influence of weather on population dynamics so that
predictions can be made about the effects of future climate
change. Several complicating factors must be confronted when
addressing these questions. First, deterministic models of population
dynamics are virtually always inadequate due to ``process variation'',
the inherent stochasticity in demographic parameters and environmental
conditions. Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of ``observation error'', such as imperfect
detection. Failure to account for  and heterogeneity in observation
error can bias the estimators of abundance and related parameters.

>>>>

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}

@@

\title{Improved state-space models for inference about spatial and
  temporal variation in abundance}

\author{Richard B. Chandler  \\
        \small USGS Patuxent Wildlife Research Center \\
        \small Laurel, MD 20708
   \and Jeffrey A. Hostetler \footnote{Contact:
     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
%        \small Migratory Bird Center \\
        \small Smithsonian Conservation Biology Institute \\
        \small National Zoological Park, MRC 5503 \\
        \small Washington, DC 20013-7012}
\maketitle

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or management actions. State-space
models are widely used for these purposes because they allow for
estimating parameters of classical growth models, such as exponential,
logistic, and Gompertz, while accounting for demographic stochasticity
and observation error. Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models for spatially-replicated
time-series data recently proposed by
\citet[Biometrics]{dail_madsen:2011}. We
expand this class of models to accommodate classical growth
models, zero-inflation, and random effects such as observer-specific
detection probabilities. Furthermore, we describe classical and
Bayesian methods of parameter estimation. These new developments will
allow researchers to apply these methods to address important
questions regarding the factors affecting spatial and temporal
variation in abundance. We demonstrate these developments by analyzing
data from the North American Breeding Bird Survey. Code to fit these
models using the \textbf{R} package \texttt{unmarked} and
\textbf{JAGS} is also provided.
}

\vspace{0.5cm}

@@

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in abundance, for
example to test for the existence of density-dependence population
growth \citep{dennis_taper:1994,viljugrein_etal:2005}. Such models are
also required in applied ecological research
for predicting species' responses to environmental change or
conservation actions. For example, %many researchers are currently
models of population dynamics allow for inference about the effects of
of climate change on population viability.
%investigations of the influence of climate on population dynamics so that
%predictions can be made about the effects of climate change.
Several complicating factors must be confronted when
addressing these questions. First, deterministic models of population
dynamics are virtually always inadequate due to process variation,
the inherent stochasticity in demographic parameters and environmental
conditions. Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection. Failure to account for process variation and observation
error can bias the estimators of abundance and related parameters.
